Why do languages tolerate heterography? An experimental investigation of the cultural evolution of informative orthography

### 2023-09-07 manuscript/wp4.tex
ms 7/09, taking in Kathy's edits

\PassOptionsToPackage{hidelinks}{hyperref}
\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\captionsetup[figure]{textfont=rm,font=small,labelfont=bf}
\usepackage{amsmath}
\usepackage{mathspec}
\setallmainfonts{Times New Roman}
\setmainfont{Times New Roman}
% \setsansfont{Helvetica Neue}

@@

We recruited 287 participants via the Prolific platform. Participants were paid £2.00 for participation plus additional bonuses of up to £1.08 as detailed below (median bonus: £0.74). The median completion time was 15~m with a median hourly rate of £8.05 (£10.94 including bonus). We limited recruitment to (self-declared) native English speakers, since it was important that participants would perceive the spoken forms in a relatively consistent way (particularly in the case of Experiment~2). 14 participants were excluded because they (or their partners) used English color words (8) or failed the auditory attention checks (6). A further three participants were lost to communication-game pairing failures. The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of subjects).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/stimuli.eps}}
	\vspace*{2pt}
	\caption{The nine alien object stimuli with their iconic stems (indicating shape) and suffixes (indicating color). The spoken forms of the suffixes are homophonous, but their associated spellings are free to evolve and may therefore take on differentiated forms.}
	\label{stimuli}
	\end{figure}

@@

Each of the 10 transmission chains was seeded with a randomly-generated suffix spelling system. This was created by randomly mapping the following nine spellings onto the nine objects: <co>, <coe>, <coh>, <ko>, <koe>, <koh>, <qo>, <qoe>, <qoh>. In other words, the /k/ sound may be spelled <c>, <k>, or <q> and the /əʊ/ sound may be spelled <o>, <oe>, or <oh>, although the initial seed system contains no particular regularity. This models an initial state of high spelling variation (every object has a unique suffix spelling), which may, over time, become transparent (the /k/ and/or /əʊ/ sounds take on consistent spellings) or expressive (spellings of /k/ and/or /əʊ/ become systematically associated with meaning). The spoken forms were synthesized using the Apple text-to-speech synthesizer (Tessa voice).

@@

As outlined above, the participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A). The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was culturally transmitted. Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings generated at generation $i-1$ (at least one of each shape and at least one of each color). Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items. Transmission continued for nine generations in each of ten independent chains. In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consists of a pair of participants, but the productions of only one of the participants (the primary participant) are iterated to the next generation (the productions of the secondary participant are thus a cultural deadend). The role of the secondary participant is to act as a communicative partner for the primary participant, inducing pressure for the language to become informative.

@@

All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials. Participants were also told explicitly that the stems were iconic of the objects' shapes, allowing participants to focus on learning the suffixes during the training phase. In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each. In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type in the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text). The participant received a 2p bonus for spelling the word correctly but had to submit their response within 20~s. Each of the six object--word pairs in the training set (i.e., the seen items that passed through the bottleneck on transmission) was passively exposed 18 times and mini-tested six times, resulting in a total of 108 passive exposure trials and 36 mini-test trials (the maximum bonus in training was therefore 72p). To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded.

@@

After training, participants assigned to the Transmission-only condition completed a test phase, alternating between production and comprehension trials. In production trials, the participant was shown an object and heard its associated word pronounced aloud. The participant's task was to type~in the appropriate spelling. The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial. Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem. This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix. Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix. In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice). In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided. Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).

@@

Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit on an independently selected set of seen items). The game closely mirrored the overall structure of the test administered to participants in the non-communicative condition, with the production and comprehension trials becoming the two sides of a single communicative interaction. On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array). The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).

@@

The results from all ten chains (labelled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}. Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape down the rows and color along the columns, just as in Fig.~\ref{stimuli}. For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension. We describe such a system as ``redundant'' because shape is always expressed by the stem, so the suffix spellings that emerged in this case therefore convey no additional information. Similar outcomes occurred in Chains~C, E, F, and~H and are characteristic of the Transmission-only condition. We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J. Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language. Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>. The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.

@@

To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate. These four typological categories may be positioned along a complexity/cost continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{typology}. The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing. To classify a participant's output into one of these four categories, we compute which of the reference systems is most similar in terms of its information content. To do this, we formalize the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite[VoI;][]{Meila:2007} defined as the distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another. Under this metric, the holistic and degenerate systems are considered very dissimilar because they carry very different levels of information content. The expressive and redundant systems are also considered very dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} Then, rather than simply classify a given system into exactly one typological category, we compute a probability distribution over the typological space according to the respective distances of the four reference systems. This gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially suggestive of one typological category but also suggestive of another (e.g., T9 is part redundant, part degenerate). The typological probability distribution $t$ over the set of four reference systems $R$ was calculated for participant system $s$ according to% $t(i) \propto e^{-2\mathrm{VoI}(R_i, s)^2}$,
\begin{equation}
	t(i) \propto e^{-2 \cdot \mathrm{VoI}(R_i, s)^2}
\end{equation}
and averaged over chains for a given generation.

@@

As a quantitative measure of how informative the written language is over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}. Communicative cost expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. Formally, it is given by% $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
	\begin{equation}
	\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
	\end{equation}
where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.

@@

Fig.~\ref{cost_dif} plots communicative cost over the nine generations of cultural transmission in each condition. In the upper panels, each thin line represents an individual transmission chain, and the thick lines shows the mean across chains. The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling. In both conditions, cost increases with generation, implying that the languages become less informative over time. This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent. In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems. This result does, however, align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produced differentiated spellings, be it through explicit innovation or implicit generalization. Interestingly, however, cost plateaus at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication, suggesting that there is some tendency for the systems to remain slightly more informative in the communicative condition.

@@

To evaluate this argument statistically, we regressed cost on generation in a Bayesian, hierarchical, polynomial regression model. The posterior parameter estimates are shown in Fig.~\ref{posterior_dif}. Essentially, the $\beta$ parameters of the model (characterizing the shape of the polynomial) do not differ substantially between conditions, but the intercepts ($\alpha$) do appear to differ. Since generation was centered (for the purpose of fitting the model), the $\alpha$ parameters correspond to Generation~5 and therefore represent the main plateaus of the polynomials. The posterior difference between $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ was 0.301~bits (95\%~HDI: 0.014--0.597). This difference, and its credible interval, is greater than zero, pointing to an effect of condition on the level at which cost plateaus. We emphasize, however, that this difference represents a relatively small effect size, corresponding to a 9.4 percentage point advantage (in terms of communicative accuracy) for systems that emerged in the communicative condition. Specifically, systems that emerged under Transmission-only would, on average, yield a communicative success rate of 40.7\%, whereas the equivalent value for Transmission + Communication is 50.1\%. Thus, although the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use.

@@

Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation. Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or arbitrary silent letters. However, such differentiation was not very forthcoming, even under communicative pressure. Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure.

@@

It is particularly surprising that participants chose not to innovate novel systems in communication, despite their being a clear financial incentive to do so. Why might this be the case? Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training. Participants who adopted this interpretation may have feared diverging from their input, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules. Secondly, even if participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners? They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems.

These are dynamics that also exist in the real world, especially in writing where the reader and writing are separated in time and space. Deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>). And even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If you decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take your spelling seriously or even understand the distinction? Differentiation is hard in the written modality.

@@

The experiment was completed by 297 native-English participants recruited through Prolific. The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.78). The median completion time was 15~m with a median hourly rate of £7.99 (£10.68 including bonus). 17 participants were excluded because they (or their partners) failed the auditory attention checks (15) or used English color words (2). A further 10 participants were lost to communication-game pairing failures. Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 subjects) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of subjects).

@@

The alien objects and word stems were identical to Experiment~1. Unlike Experiment~1, however, the transmission chains were seeded with a fully compositional language that uses three distinct suffixes to systematically express each of the colors. A separate set of suffixes was created for each chain by concatenating a randomly drawn consonant from \{/f/, /s/, /ʃ/\} and a randomly drawn vowel from \{/ə/, /ɛɪ/, /əʊ/\}, both without replacement. For example, one chain might use the suffixes /fəʊ/, /ʃə/, /sɛɪ/, while another chain might use /sə/, /fɛɪ/, /ʃəʊ/. The initial orthographic system was transparent and based on the following phoneme–grapheme mapping: \{/f/→<f>, /s/→<s>, /ʃ/→<x>, /ə/→<a>, /ɛɪ/→<ei>, /əʊ/→<oe>\}. The suffixes were designed to be distinctive (and therefore easy to memorize and associate with colors), but also similar enough to (mostly) allow for somewhat plausible sound mergers and result in somewhat plausible spellings following sound merger (e.g., it is plausible that /f/ might supplant /s/ or /ʃ/ in speech or that /ʃ/ might be spelled <s> or <x> in writing). We attempted to achieve this balance by combining consonants that are very similar with vowels that are very dissimilar, while also avoiding reuse of any sounds present in the stems. The spoken forms were synthesized using the Apple text-to-speech synthesizer (Moira voice).

@@

Fig.~\ref{cost_con} plots communicative cost. These mean lines reveal a fairly clear effect from epoch. During Epoch~I, there is a shallow increase in cost with generation in both conditions. This represents a kind of baseline level of noise when the spoken languages are fully informative. During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in Transmission + Communication. This is because the communicative pressure acts to slow down the rate at which the systems become transparent or redundant. In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change. In Transmission + Communication, by contrast, the curve gets a little steeper because the systems still have some room to further simplify and the pressure to simplify has been ramped up.

>>>>

\PassOptionsToPackage{hidelinks}{hyperref}
\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\captionsetup[figure]{textfont=rm,font=small,labelfont=bf}
\usepackage{amsmath}
\usepackage{mathspec}
\setallmainfonts{Times New Roman}
\setmainfont{Times New Roman}

@@

We recruited 287 participants via the Prolific platform. Participants were paid £2.00 for participation plus additional bonuses of up to £1.08 as detailed below (median bonus: £0.74). The median completion time was 15~m with a median hourly rate of £8.05 (£10.94 including bonus). We limited recruitment to (self-declared) native English speakers, since it was important that participants would perceive the spoken forms in a relatively consistent way (particularly in the case of Experiment~2). 14 participants were excluded because they (or their communication partners) used English color words (8) or failed the auditory attention checks (6). A further three participants were lost to communication-game pairing failures. The final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 participants) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of participants).

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/stimuli.eps}}
	\vspace*{2pt}
	\caption{The nine alien object stimuli with their stems and suffixes. The spoken and written forms of the stems were fixed and unchanging throughout the experiment, as were the spoken forms of the suffixes, which were always homophonous; however, the written forms of the suffixes were free to evolve over time, potentially taking on differentiated forms to indicate color (e.g., <-co>, <-ko>, and <-qo> to represent pink, yellow, and blue).}
	\label{stimuli}
	\end{figure}

@@

Each of the 10 transmission chains was seeded with a randomly-generated suffix spelling system, which was created by randomly mapping the following nine spellings onto the nine objects: <co>, <coe>, <coh>, <ko>, <koe>, <koh>, <qo>, <qoe>, <qoh>. In other words, the /k/ sound may be spelled <c>, <k>, or <q> and the /əʊ/ sound may be spelled <o>, <oe>, or <oh>, although the initial seed system contains no particular regularity. This procedure models an initial state of high spelling variation (every object has a unique suffix spelling), but these spellings may, over time, become transparent (the /k/ and/or /əʊ/ sounds take on consistent spellings) or expressive (spellings of /k/ and/or /əʊ/ become systematically associated with meaning). The spoken forms were synthesized using the Apple text-to-speech synthesizer (Tessa voice).

@@

Participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A). The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was transmitted to subsequent generations. Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings produced at generation $i-1$ (at least one of each shape and at least one of each color). Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items. Transmission continued for nine generations in each of ten independent chains. In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consists of a pair of participants, but the productions of only one of the participants (the primary participant; determined by whichever participant started the experiment first) are iterated to the next generation (the productions of the secondary participant are thus a cultural deadend). The role of the secondary participant is to act as a communicative partner for the primary participant, inducing pressure for the language to become informative.

@@

All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials. Participants were also told explicitly that the stems were iconic of the objects' shapes, allowing participants to focus on learning the suffixes during the training phase. In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each. In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text). The participant received a 2p bonus for spelling the word correctly but had to submit their response within 20~s. Each of the six object--word pairs in the training set (i.e., the seen items that passed through the bottleneck on transmission) was passively exposed 18 times and mini-tested six times, resulting in a total of 108 passive exposure trials and 36 mini-test trials (the maximum bonus in training was therefore 72p). To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded.

@@

After training, participants assigned to the Transmission-only condition completed a test phase, alternating between production and comprehension trials. In production trials, the participant was shown an object and heard its associated pronunciation. The participant's task was to type the appropriate spelling. The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial. Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem. This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix. Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix. In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice). In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided. Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).

@@

Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit subject to a separate application of the bottleneck on transmission; i.e., the seen items were selected independently for each participant). The game closely mirrored the overall structure of the test administered to participants in the non-communicative condition, with the production and comprehension trials becoming the two sides of a single communicative interaction. On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array). The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).

@@

The results from all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}. Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, just as in Fig.~\ref{stimuli}. For example, the system at Generation~9 in Chain~D uses three spellings---<koe>, <ko>, and <co>---to express the shape dimension. We describe such a system as ``redundant'' because shape is always expressed by the stem, so the suffix spellings that emerged in this case therefore convey no additional information. Similar outcomes occurred in Chains~C, E, F, and~H and are characteristic of the Transmission-only condition. We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J. Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language. Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>. The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.

@@

To analyze more formally what types of system tend to emerge under different conditions, we first designated four reference systems that are of particular interest: holistic,\footnote{Although we use the term holistic here, this category also encompasses compositional suffixes that express both shape and color in a systematic way (specifically, K9, M1, M7, and Q7).} expressive, redundant, and degenerate. These four typological categories may be positioned along a complexity/cost continuum in terms of the number of suffix forms they make use of and how these forms are conditioned on meaning, as illustrated in Fig.~\ref{typology}. The holistic system has nine unique forms that express both color and shape, the expressive and redundant systems have three unique forms that express either color (expressive) or shape (redundant), and the degenerate category has one unique form that expresses nothing. To classify a participant's output into one of these four categories, we may compute which of the reference systems is most similar in terms of its information content. To do this, we formalized the systems as set partitions (i.e., a partitioning of the universe of meanings into disjoint subsets) with variation of information \parencite[VoI;][]{Meila:2007} defined as the distance metric between any two such partitions.\footnote{VoI is a proper metric on set partitions, measuring the amount of information (in bits) that is lost and gained in the transformation of one set partition into another. Under this metric, the holistic and degenerate systems are considered very dissimilar because they carry very different levels of information content. The expressive and redundant systems are also considered quite dissimilar because, although they carry the same amount of information, the information they carry is orthogonal (shape in the case of redundant, and color in the case of expressive).} Then, rather than simply classify a given system into exactly one typological category, we computed a probability distribution over the typological space according to the respective distances from the four reference systems. This approach gives a slightly more nuanced view of the typological distribution by accounting for cases where a given system might be partially suggestive of one typological category but also suggestive of another (e.g., T9 is part redundant, part degenerate). The typological probability distribution $t$ over the set of four reference systems $R$ was calculated for participant system $s$ according to% $t(i) \propto e^{-2\mathrm{VoI}(R_i, s)^2}$,
\begin{equation}
	t(i) \propto e^{-2 \cdot \mathrm{VoI}(R_i, s)^2}
\end{equation}
and averaged over chains for a given generation.

@@

As a quantitative measure of how informative the written language is over time, we adopt communicative cost \parencite{KempRegier:2012, Regier:2015, Kemp:2018}, which expresses how much information is lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. Formally, it is given by% $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$,
	\begin{equation}
	\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) ),
	\end{equation}
where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system.

@@

Fig.~\ref{cost_dif} plots communicative cost over the nine generations of cultural transmission in each condition. In the upper panels, each thin line represents an individual transmission chain, and the thick lines show the mean across chains. The initial, generation-0 seed systems are all fully informative (zero cost) because each meaning is signaled by a unique arbitrary spelling. In both conditions, cost increases with generation, implying that the languages become less informative over time. This is to be expected in the Transmission-only condition, since there is only pressure from learnability, so the systems evolve to become simple and transparent. In contrast, this is not the expected result in Transmission + Communication, where the communicative pressure should promote the emergence of informative systems. This result does, however, align with our qualitative interpretation of the data, which suggested that participants in the communicative condition rarely produced differentiated spellings, be it through explicit innovation or implicit generalization. Interestingly, however, cost plateaus at around 1.3~bits in Transmission-only and around 1~bit in Transmission + Communication, suggesting that there is some tendency for the systems to remain slightly more informative in the communicative condition.

@@

To evaluate this argument statistically, we regressed cost on generation in a Bayesian, hierarchical, polynomial regression model. 
    \begin{align*}
       c_i & \sim \mathrm{Normal}(\mu_i, \sigma) \\
     \mu_i & =    \alpha_{\mathrm{CHAIN}[i]} + \beta_{\mathrm{CHAIN}[i]} g_i + \gamma_{\mathrm{CHAIN}[i]} g_i^2 + \delta_{\mathrm{CHAIN}[i]} g_i^3 \\
    \sigma & \sim \mathrm{Exponential}(1)
    \end{align*}
The posterior parameter estimates are shown in Fig.~\ref{posterior_dif}. Essentially, the $\beta$ parameters of the model (characterizing the shape of the polynomial) do not differ substantially between conditions, but the intercepts ($\alpha$) do appear to differ. Since generation was centered (for the purpose of fitting the model), the $\alpha$ parameters correspond to Generation~5 and therefore represent the main plateaus of the polynomials. The posterior difference between $\alpha_\mathrm{tran}$ and $\alpha_\mathrm{comm}$ was 0.301~bits (95\%~HDI: 0.014--0.597). This difference, and its credible interval, is greater than zero, pointing to an effect of condition on the level at which cost plateaus. We emphasize, however, that this difference represents a relatively small effect size, corresponding to a 9.4 percentage point advantage (in terms of communicative accuracy) for systems that emerged in the communicative condition. Specifically, systems that emerged under Transmission-only would, on average, yield a communicative success rate of 40.7\%, whereas the equivalent value for Transmission + Communication is 50.1\%. Thus, although the presence of the communicative pressure does appear to make a difference to the informativeness of the emergent systems, that difference is not especially impressive considering that an ideal, expressive system of suffix spellings would result in zero cost and a 100\% success rate in communicative use.

@@

Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation. Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <a> for pink, <b> for yellow, and <c> for blue). However, such differentiation was not very forthcoming, even under communicative pressure. Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure.

@@

It is particularly surprising that participants chose not to innovate novel systems in communication, despite their being a clear financial incentive to do so. Why might this be the case? Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training. Participants who adopted this interpretation may have feared diverging from what they learned, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules. Secondly, even if participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners? They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems.

These are dynamics that also exist in the real world, especially in writing where the reader and writer are separated in time and space. Deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>). Even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the distinction? Differentiation is hard in the written modality.

@@

The experiment was completed by 297 native-English participants recruited through Prolific. The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.78). The median completion time was 15~m with a median hourly rate of £7.99 (£10.68 including bonus). 17 participants were excluded because they (or their partners) failed the auditory attention checks (15) or used English color words (2). A further 10 participants were lost to communication-game pairing failures. Like Experiment~1, the final dataset comprises 270 participants: 90 in the Transmission-only condition (10 chains of 9 participants) and 180 in the Transmission~+ Communication condition (10 chains of 9 pairs of participants).

@@

The alien objects and word stems were identical to Experiment~1. Unlike Experiment~1, however, the transmission chains were seeded with a fully compositional language that used three distinct suffixes to systematically express each of the colors. A separate set of suffixes was created for each chain by concatenating a randomly drawn consonant from \{/f/, /s/, /ʃ/\} and a randomly drawn vowel from \{/ə/, /ɛɪ/, /əʊ/\}, both without replacement. For example, one chain might use the suffixes /fəʊ/, /ʃə/, /sɛɪ/ to represent the colors pink, yellow, and blue, while another chain might use /sə/, /fɛɪ/, /ʃəʊ/ for those colors. The initial orthographic system was transparent and based on the following phoneme–grapheme mapping: \{/f/→<f>, /s/→<s>, /ʃ/→<x>, /ə/→<a>, /ɛɪ/→<ei>, /əʊ/→<oe>\}. The suffixes were designed to be distinctive (and therefore easy to memorize and associate with colors), but also similar enough to (mostly) allow for somewhat plausible sound mergers and result in somewhat plausible spellings following sound merger (e.g., it is plausible that /f/ might supplant /s/ or /ʃ/ in speech or that /ʃ/ might be spelled <s> or <x> in writing). We attempted to achieve this balance by combining consonants that are very similar with vowels that are very dissimilar, while also avoiding reuse of any sounds present in the stems. The spoken forms were synthesized using the Apple text-to-speech synthesizer (Moira voice).

@@

Fig.~\ref{cost_con} plots communicative cost. The mean lines reveal a fairly clear effect of epoch. During Epoch~I, there is a shallow increase in cost with generation in both conditions. This represents a baseline level of noise when the spoken languages are fully informative. During Epoch~II, there is a steep rise in cost in the Transmission-only condition, but a relatively shallow rise in the Transmission + Communication condition. This difference is presumably due to the communicative pressure, which acts to slow down the rate at which the systems become transparent or redundant. In Epoch~III, the cost curve plateaus in the Transmission-only condition because the written languages have become sufficiently simple to be reliably learned and reproduced with little change. In Transmission + Communication, by contrast, the curve gets a little steeper because the systems still have some room to further simplify and the pressure to simplify has been ramped up.
